The markings of the trade: exploring the patterns of olive oil production in Roman Baetica

### 2020-05-12 Paper/jaccardbaetica.tex
second draft finish

\newcommand{\memo}[2]{\textcolor{#1}{#2}}
\newcommand{\maria}[1]{\memo{red}{#1\\}}
\newcommand{\xavi}[1]{\memo{magenta}{XRC: #1\\}}

@@

\begin{keyword}
Roman Empire; amphora workshops; Dressel 20; social learning; cultural evolution
\end{keyword}

@@

The aim of this study is to explore economics dynamics in the production and distribution of olive oil trade. 
Our case of study has been focused on the production processes located in \textit{Baetica} province (currently Andalusia) from 1st to 3rd AD. In particular, we want to detect patterns of olive oil production that link amphora workshops and amphoric stamps. \textit{Baetica} became important production and distribution centre during the Roman Empire. However, it remains under debate about how this province was organised and whether it could be possible to identify patterns in the olive oil market. Amphoric stamps are used to identify the presence of different groups that might share similar stamps. To achieve this goal, we analyse a set of stamps from two centres: 1) producer centres by analysing different workshops in \textit{Baetica} province and 2) two Roman provinces such as \textit{Germania} and \textit{Britannia} as a receptor centres. They will be used to detect a connection between the distribution of amphoric stamps and the economic structure in both centres. Here we use methods borrowed from Ecology that allow us to identify if amphora workshops share similar amphoric stamps depending on the spatial distance.

@@

Given this fact, an important mechanism of production control under the Roman government was created spreading by different areas attending to the richness of different places. Roman provinces such as \textit{Baetica} (currently Andalusia, south Spain) became important production centres of olive oil during the Roman Empire. Olive oil was considered the liquid gold since it was used in different aspect of daily life for cooking, hygiene or lighting  \citep{mattingly_d.j._oil_1988}. As a consequence, this product was distributed in a large number of amphorae along the provinces, mostly to supply the Roman Army and Italy \citep{blazquez_exportacion_1980}.

@@

The location of the workshops with a riverine connectivity allowed to set up a connection with the Mediterranean and Atlantic routes \citep{garcia_vargas_enrique_formal_2010}.

@@

The majority of olive oil amphorae produced in this province belong to Dressel 20 type\citep{dressel_ricerche_1878,
millet_anforas_1998}. Dressel 20 amphorae were shipped and distributed from Baetica throughout western Europe.

Dressel 20 contains many information still unknown. A high percentage of  Dressel 20 were marked in stamps, inked in \textit{tituli picti} and incise in \textit{graffiti}. Indeed, most of the inscriptions on the amphora have not enough analysed due to the fragmentation of the material or the shortage of samples \citep{aguilera_evolucion_2007,rovira_guardiola_grafitos_2007}.

@@

Most stamps display a large activity of production that it can be difficult for specifying an accurate chronology in some cases. They have been frequently dated with the consultar year by studying \textit{tituli picti} found in \textit{Monte Testaccio} \citep{Testaccio1, berni_millet_epigrafianforica_2008}.
However, the chronology of stamps by using \textit{tituli picti} could be also biased in other sites taking into account that amphorae were deposited after being produced and consumed.  
\maria{no sé cómo explicarlo mejor}

The fascination aroused by stamps has caused it to be one of the inscriptions with the most publications about its origin and meaning. Its endure in time become an essential element that were used to mark amphorae during almost three centuries \citep{remesal_sellar_2016} (Fig.\ref{amphora}).

@@

They could be identified based on three premises: content (olive oil), context (amphora workshop) and subject (individuals involved). On the one hand, it seems that stamps could have been identified as the land owner of the olive groves \citep{rodriguez_economioleicola_1977}. On the other, they could belong to the owner of the making amphorae workshops or even a group of amphora workers \citep{berni_millet_epigrafianforica_2008}. In any case, the use of these stamps became in a good proxy to define somehow the system of working in the workshops.

@@

\section{Case study: consumption centers}

The conquest of new provinces allowed to the Roman Empire the arrival of new resources through Mediterranean and Atlantic routes. This lead to a progressive change in the economic and social structure creating new trade networks for supplying Roman conquest. As a result, Augusto's administration created the figure of the \textit{praefectura annonae} for the supply of wheat. The role of the \textit{praefectura annonae} was mainly to provide though \textit{frumentationes} a quantity per month of wheat to Roman citizens \citep{remesal_annona_1986,remesal_concierto}.

@@

The importance of olive oil production could be explained by the fact that we detected a massive amphora distribution in different Roman provinces such as \textit{Germania} and \textit{Britannia}. This amphora has been commonly associated with the  transportation of Baetican olive oil for supplying military camps and civil settlements \citep{berni_millet_epigrafianforica_2008}.

The concentration of amphorae in eminently militarised areas suggests a possible existence of a broad supply mechanism, being accurate to imagine a possible organisation in charge of its distribution from military camps to civil settlements\citep{remesal_annona_1986, carreras_britannia_1998}.

The military consumption of olive oil in Roman provinces could be related to two aspects: a) cultural consumption whose product is consumed by cultural reasons such as to identity or habit and b) economical reasons where olive oil distribution is based on the transport costs unlike other products \citep[69-70]{carreras_britannia_1998}.

\maria{falta conexión entre una cosa y hacer una breve introducción de lo siguiente britannia y germania}

@@

\xavi{en general, echo a faltar un poco más de detalle en esta discusión y en los datos que tienes; en qué se parecen y diferencian las 2 provincias? Si hablas del uso por las campañas o las guarniciones, donde están? Qué strategic points de redistribución hay? Tampoco a saco, pero entender mejor cómo llega el aceite, cómo se redistribuye y quién y dónde se consume}

@@

\xavi{Esta estructura no me convence...tenemos que pensar algo mejor, porque ahora es: 1. hablo de britania/germania, 2. hablo de sellos en general, 3. hablo de sellos en britania/germania. Quizás juntar esto con lo de los casos de estudio, o quizás hablar de los 2 casos de estudio en la intro, y dividir la sección anterior en 2 (lo más general a la intro, lo más concreto aquí)}

@@

The analysis of the amphora workshops in Baetica province resulted in a significant correlation between spatial distance and the similarity of the amphoric stamps.

>>>>

\newcommand{\memo}[2]{\textcolor{#1}{#2}}
\newcommand{\maria}[1]{\memo{red}{MC: #1\\}}
\newcommand{\xavi}[1]{\memo{magenta}{XRC: #1\\}}

@@

\begin{keyword}
Roman Empire; amphora production; Dressel 20; dissimilarity index; Roman provinces
\end{keyword}

@@

The aim of this study is to explore economics dynamics in the production and distribution of olive oil trade. 
Our case of study has been focused on the production processes located in \textit{Baetica} province (currently Andalusia) from 1st to 3rd AD. In particular, we want to detect patterns of olive oil production that link amphora workshops and amphoric stamps. \textit{Baetica} became an important production and distribution centre during the Roman Empire. However, it remains under debate about how this province was organised and whether it could be possible to identify patterns in the olive oil market. Amphoric stamps are used to identify the presence of different groups that might share similar stamps. To achieve this goal, we analyse a set of stamps from two centres: 1) producer centres by analysing different workshops in \textit{Baetica} province and 2) two Roman provinces such as \textit{Germania} and \textit{Britannia} as a receptor centres. They will be used to detect a connection between the distribution of amphoric stamps and the economic structure in both centres. Here, we use methods borrowed from Ecology that allow us to identify if amphora workshops share similar amphoric stamps depending on the spatial distance.

@@

Given this fact, an important mechanism of production control under the Roman government was created spreading by different areas attending to the richness of different places. Roman provinces such as \textit{Baetica} (currently Andalusia, south Spain) became important production centres of olive oil during the Roman Empire. Olive oil was considered the liquid gold since it was used in different aspect of daily life for cooking, hygiene or lighting  \citep{mattingly_d.j._oil_1988}. As a consequence, this product was distributed in a large number of amphorae across the provinces, mostly to supply the Roman Army and Italy \citep{blazquez_exportacion_1980}.

@@

The location of the workshops with a riverine connectivity allowed to arrange a connection with the Mediterranean and Atlantic routes \citep{garcia_vargas_enrique_formal_2010}.

@@

The majority of olive oil amphorae produced in this province belong to Dressel 20 type\citep{dressel_ricerche_1878,
millet_anforas_1998}. Dressel 20 amphorae were shipped and distributed from \textit{Baetica} throughout western Europe.

Dressel 20 contains much information still unknown. A high percentage of  Dressel 20 were marked in stamps, inked in \textit{tituli picti} and incise in \textit{graffiti}. Indeed, most of the inscriptions on the amphora have not enough analysed due to the fragmentation of the material or the shortage of samples \citep{aguilera_evolucion_2007,rovira_guardiola_grafitos_2007}.

@@

Most stamps display a large activity of production that it can be difficult for specifying an accurate chronology in some cases. They have been frequently dated with the consular dating by studying \textit{tituli picti} found in \textit{Monte Testaccio} \citep{Testaccio1, berni_millet_epigrafianforica_2008}.
However, the chronology of stamps by using \textit{tituli picti} could be also biased in other sites taking into account that amphorae were deposited after being produced and consumed.  
\maria{no sé cómo explicarlo mejor ni como citarlo porque no hay citas}

The fascination aroused by stamps has caused it to be one of the inscriptions with the most publications about its origin and meaning. Its endure in time become an essential element that was used to mark amphorae during almost three centuries \citep{remesal_sellar_2016} (Fig.\ref{amphora}).

@@

They could be identified based on three premises: content (olive oil), context (amphora workshop) and subject (individuals involved). On the one hand, it seems that stamps could have been identified as the landowner of the olive groves \citep{rodriguez_economioleicola_1977}. On the other, they could belong to the owner of the making amphorae workshops or even a group of amphora workers \citep{berni_millet_epigrafianforica_2008}. In any case, the use of these stamps became in a good proxy to define somehow the system of working in the workshops.

@@

\section{Case study: consumption centres}

The conquest of new provinces allowed the Roman Empire the arrival of new resources through Mediterranean and Atlantic routes. This lead to a progressive change in the economic and social structure creating new trade networks for supplying Roman conquest. As a result, Augusto's administration created the figure of the \textit{praefectura annonae} for the supply of wheat. The role of the \textit{praefectura annonae} was mainly to provide though \textit{frumentationes} a quantity per month of wheat to Roman citizens \citep{remesal_annona_1986,remesal_concierto}.

@@

The importance of olive oil production could be explained by the fact that we detected a massive amphora distribution in different Roman provinces. This amphora has been commonly associated with the  transportation of Baetican olive oil for supplying military camps and civil settlements \citep{berni_millet_epigrafianforica_2008}.

The military consumption of olive oil in Roman provinces could be related to two aspects: a) cultural consumption whose product is consumed by cultural reasons such as to identity or habit and b) economical reasons where olive oil distribution is based on the transport costs, unlike other products \citep[69-70]{carreras_britannia_1998}.

Olive oil supply had a special impact in militarised provinces such as \textit{Britannia} and \textit{Germania}. The concentration of amphorae in eminently militarised areas suggests the existence of a broad supply mechanism, being accurate to imagine a possible organisation in charge of its distribution from military camps to civil settlements\citep{remesal_annona_1986, carreras_britannia_1998}.

@@

%\xavi{en general, echo a faltar un poco más de detalle en esta discusión y en los datos que tienes; en qué se parecen y diferencian las 2 provincias? Si hablas del uso por las campañas o las guarniciones, donde están? Qué strategic points de redistribución hay? Tampoco a saco, pero entender mejor cómo llega el aceite, cómo se redistribuye y quién y dónde se consume}

@@

%\xavi{Esta estructura no me convence...tenemos que pensar algo mejor, porque ahora es: 1. hablo de britania/germania, 2. hablo de sellos en general, 3. hablo de sellos en britania/germania. Quizás juntar esto con lo de los casos de estudio, o quizás hablar de los 2 casos de estudio en la intro, y dividir la sección anterior en 2 (lo más general a la intro, lo más concreto aquí)}

@@

The analysis of the amphora workshops in \textit{Baetica} province resulted in a significant correlation between spatial distance and the similarity of the amphoric stamps.
